---
title: "Seasonal_TaxonomyPerSeason"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

After plotting the total taxonomic composition, we now want to find out which groups (orders or families or genera) dominate which habitat. Here we will plot the taxonomic composition partitioned into the habitats in a stacked bar chart.

## Prepare Data

```{r Load Data, message=FALSE, results='hide', warning=FALSE}
rm(list = ls())
library(ggplot2)
library(plyr)
library(ggpubr)
library(reshape2)
library(viridis)

TAX = read.csv("../00_Data/04_Oomycota_Seasonal_OTU_ContingencyTable_filtered_sequences_NCBI-nt_blasted_sorted_BestHit_TaxonomyTable_Oomycetes_sequences_vsearch-ITS1-BestHit_AnnotationRefined_noPipe_withFunction_noNA.tsv", 
                    header = F, 
                    sep = "\t", 
                    stringsAsFactors = T)
OTU_Table = as.data.frame(read.csv("../00_Data/05_Oomycota_Seasonal_OTU_Table_min-freq-1172_transposed_withMetadata.tsv", 
                     header = T, 
                     sep = "\t", 
                     stringsAsFactors = T))
SampleMetadata = OTU_Table[,1:17]
OTU_Table = OTU_Table[,18:ncol(OTU_Table)]
Abundances = colSums(OTU_Table)
TAX = cbind(TAX, Abundances)
colnames(TAX) = c("OTU_Number", "Order", "Family", "Genus", "Species", "ReferenceID", "PercentID", "Lifestyle", "Substrate", "Abundance")
TAX$OTU_ID = paste0("OTU", TAX$OTU_Number, "_", TAX$Species)
TAX$Lifestyle = as.factor(TAX$Lifestyle)
TAX$Class = "Oomycota"
TAX$Order = gsub("Order_NoHit", "Undetermined", 
                           TAX$Order)
TAX$Order = gsub("incertae_sedis", "Incertae sedis", 
                           TAX$Order)

```

## Aggregate Table

Next we aggregate the table by microhabitat and order.
What we then have is a 9x8 dataframe with 9 microhabitats and 8 orders

```{r Aggregate Table}
OrderDataFunction = function(MicrohabitatPerSampling, incidence){
OrderTable = OTU_Table
colnames(OrderTable) = TAX$Order
# First aggregate by Habitat
HabitatAggregatedOrderTable = 
  aggregate(OrderTable, 
            by = list(SampleMetadata$MicrohabitatPerSampling), 
            FUN = sum)
rownames(HabitatAggregatedOrderTable) = 
  HabitatAggregatedOrderTable$Group.1
HabitatAggregatedOrderTable = 
  HabitatAggregatedOrderTable[,-1]
# if you want to check the number of OTUs instead of the amount of reads:
# this is the point to convert it into a presence/absence matrix like this:
if(incidence == T){
HabitatAggregatedOrderTable[,] = 
  ifelse(HabitatAggregatedOrderTable[,] > 0, 1, 0)
}
#Then aggregate the (transposed) table by Order
HabitatAggregatedOrderTable = 
  aggregate(t(HabitatAggregatedOrderTable), 
            by = list(TAX$Order), 
            FUN = sum)
rownames(HabitatAggregatedOrderTable) = 
  HabitatAggregatedOrderTable$Group.1
HabitatAggregatedOrderTable = 
  HabitatAggregatedOrderTable[,-1]
HabitatAggregatedOrderTable = 
  as.data.frame(HabitatAggregatedOrderTable)
# melt the table so ggplot can interpret it as a histogram
data = HabitatAggregatedOrderTable
data$Order = rownames(data)
data2 = melt(data, id.vars = "Order")
# add the Stratum
data3 = data2
data3$Stratum = ifelse(grepl("Leaf Litter|^Soil", data2$variable), 
                       "Ground", "Canopy")
# for the plotting we need to change the column names
colnames(data3) = c("Order", "Microhabitat", "Abundance", "Stratum")

data4 = data3[grepl(MicrohabitatPerSampling, data3$Microhabitat),]

data4$Microhabitat = sub(paste0(" ", MicrohabitatPerSampling), "", data4$Microhabitat)


return(data4)
}
```

## Plot stacked bar chart

Now we plot it with `ggplot` and use the Stratum as a facet:

```{r Plot Barchart, fig.width=6.2,fig.height=3,dpi=300}

plot_OrderData = function(OrderData){
g = ggplot(OrderData, aes(x = Microhabitat, y = Abundance, fill = Order)) + 
  geom_bar(position = position_fill(), stat = "identity") + 
  facet_grid(cols = vars(Stratum), scales = "free", space = "free") + 
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Relative\nAbundance") + 
  scale_fill_viridis(discrete = T, option = "B") +
  theme_minimal() +
  theme(axis.text=element_text(size=12), 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.title=element_text(size=14, face = "bold"), 
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5), 
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14, face = "bold"), 
        strip.text = element_text(size=12, face = "bold"), 
        panel.grid = element_blank())
return(g)
}

OrderAutumn2017 = OrderDataFunction("Autumn 2017", incidence = F)
Plot_Autumn2017 = plot_OrderData(OrderAutumn2017)

OrderSpring2018 = OrderDataFunction("Spring 2018", incidence = F)
Plot_Spring2018 = plot_OrderData(OrderSpring2018)

OrderAutumn2018 = OrderDataFunction("Autumn 2018", incidence = F)
Plot_Autumn2018 = plot_OrderData(OrderAutumn2018)

OrderSpring2019 = OrderDataFunction("Spring 2019", incidence = F)
Plot_Spring2019 = plot_OrderData(OrderSpring2019)

combi = ggarrange(Plot_Autumn2017, Plot_Spring2018, 
                  Plot_Autumn2018, Plot_Spring2019, 
                  labels = c("AUTO"), 
                  ncol = 2, nrow = 2, 
                  common.legend = T, legend = "right", 
                  align = "hv", vjust = 2.5)

OrderAutumn2017$Sampling = "Autumn 2017"
OrderSpring2018$Sampling = "Spring 2018"
OrderAutumn2018$Sampling = "Autumn 2018"
OrderSpring2019$Sampling = "Spring 2019"

OrderDataAll = rbind(OrderAutumn2017, OrderSpring2018, 
                     OrderAutumn2018, OrderSpring2019)
OrderDataAll$Microhabitat = factor(OrderDataAll$Microhabitat, 
                                   levels = c("Arboreal Soil", "Bark", 
                                              "Deadwood", "Fresh Leaves", 
                                              "Hypnum", "Lichen", 
                                              "Orthotrichum", 
                                              "Leaf Litter", "Soil"))

g = ggplot(OrderDataAll, aes(x = Microhabitat, y = Abundance, fill = Order)) + 
  geom_bar(position = position_fill(), stat = "identity") + 
  facet_grid(cols = vars(Stratum), scales = "free", space = "free") + 
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Relative\nAbundance") + 
  scale_fill_viridis(discrete = T, option = "B") +
  theme_minimal() +
  geom_vline(xintercept = 7.5) +
  theme(axis.text=element_text(size=12), 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.title=element_text(size=14, face = "bold"), 
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5), 
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14, face = "bold"), 
        strip.text = element_text(size=12, face = "bold"), 
        panel.grid = element_blank()) + 
  facet_wrap(facets = Sampling~.)

ggsave("TaxonomyPerSeason.png", plot = g, 
       device = "png", dpi = 300, width = 18, height = 12, 
       units = "cm")
ggsave("TaxonomyPerSeason.jpeg", plot = g, 
       device = "jpeg", dpi = 300, width = 18, height = 12, 
       units = "cm")
ggsave("TaxonomyPerSeason.pdf", plot = g, 
       device = "pdf", dpi = 300, width = 18, height = 12, 
       units = "cm")

g
```

## Genus Data

Here we update the plot to highlight the comparatively high proportion of OTUs assigned to the Genus Hyaloperonospora. To do so, we add grey squares on the plot indicating the Abundance of Hyaloperonospora: 

```{r Genus Data}
GenusDataFunction = function(MicrohabitatPerSampling, incidence){
GenusTable = OTU_Table
colnames(GenusTable) = TAX$Genus
# First aggregate by Habitat
HabitatAggregatedGenusTable = 
  aggregate(GenusTable, 
            by = list(SampleMetadata$MicrohabitatPerSampling), 
            FUN = sum)
rownames(HabitatAggregatedGenusTable) = 
  HabitatAggregatedGenusTable$Group.1
HabitatAggregatedGenusTable = 
  HabitatAggregatedGenusTable[,-1]
# if you want to check the number of OTUs instead of the amount of reads:
# this is the point to convert it into a presence/absence matrix like this:
if(incidence == T){
HabitatAggregatedGenusTable[,] = 
  ifelse(HabitatAggregatedGenusTable[,] > 0, 1, 0)
}
#Then aggregate the (transposed) table by Genus
HabitatAggregatedGenusTable = 
  aggregate(t(HabitatAggregatedGenusTable), 
            by = list(TAX$Genus), 
            FUN = sum)
rownames(HabitatAggregatedGenusTable) = 
  HabitatAggregatedGenusTable$Group.1
HabitatAggregatedGenusTable = 
  HabitatAggregatedGenusTable[,-1]
HabitatAggregatedGenusTable = 
  as.data.frame(HabitatAggregatedGenusTable)
# melt the table so ggplot can interpret it as a histogram
data = HabitatAggregatedGenusTable
data$Genus = rownames(data)
data2 = melt(data, id.vars = "Genus")
# add the Stratum
data3 = data2
data3$Stratum = ifelse(grepl("Leaf Litter|^Soil", data2$variable), 
                       "Ground", "Canopy")
# for the plotting we need to change the column names
colnames(data3) = c("Genus", "Microhabitat", "Abundance", "Stratum")

data4 = data3[grepl(MicrohabitatPerSampling, data3$Microhabitat),]

data4$Microhabitat = sub(paste0(" ", MicrohabitatPerSampling), "", data4$Microhabitat)


return(data4)
}

GenusAutumn2017 = GenusDataFunction("Autumn 2017", incidence = F)
GenusSpring2018 = GenusDataFunction("Spring 2018", incidence = F)
GenusAutumn2018 = GenusDataFunction("Autumn 2018", incidence = F)
GenusSpring2019 = GenusDataFunction("Spring 2019", incidence = F)

GenusAutumn2017$Sampling = "Autumn 2017"
GenusSpring2018$Sampling = "Spring 2018"
GenusAutumn2018$Sampling = "Autumn 2018"
GenusSpring2019$Sampling = "Spring 2019"

GenusDataAll = rbind(GenusAutumn2017, GenusSpring2018, 
                     GenusAutumn2018, GenusSpring2019)
GenusDataAll$Microhabitat = factor(GenusDataAll$Microhabitat, 
                                   levels = c("Arboreal Soil", "Bark", 
                                              "Deadwood", "Fresh Leaves", 
                                              "Hypnum", "Lichen", 
                                              "Orthotrichum", 
                                              "Leaf Litter", "Soil"))

Ordervector = vector()
for(Genus in GenusDataAll$Genus){
  a = TAX[TAX$Genus == Genus, "Order"][1]
  Ordervector = c(Ordervector, a)
}

GenusDataAll$Order = Ordervector
GenusDataAll$Hyaloperonospora = ifelse(GenusDataAll$Genus == "Hyaloperonospora", 
                                       "Hyaloperonospora", NA_character_)

g_Genus = ggplot(GenusDataAll, aes(x = Microhabitat, y = Abundance, fill = Order, color = Hyaloperonospora)) + 
  geom_bar(position = position_fill(), stat = "identity", size = 0.5) + 
  facet_grid(cols = vars(Stratum), scales = "free", space = "free") + 
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Relative\nAbundance") + 
  scale_fill_viridis(discrete = T, option = "B") +
  theme_minimal() +
  geom_vline(xintercept = 7.5) +
  theme(axis.text=element_text(size=12), 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.title=element_text(size=14, face = "bold"), 
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5), 
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14, face = "bold"), 
        strip.text = element_text(size=12, face = "bold"), 
        panel.grid = element_blank()) + 
  facet_wrap(facets = Sampling~.) + 
  scale_color_manual(values = c('grey', NA), guide=F)

ggsave("TaxonomyPerSeason_revised.png", plot = g_Genus, 
       device = "png", dpi = 300, width = 18, height = 12, 
       units = "cm")
ggsave("TaxonomyPerSeason_revised.jpeg", plot = g_Genus, 
       device = "jpeg", dpi = 300, width = 18, height = 12, 
       units = "cm")
ggsave("TaxonomyPerSeason_revised.pdf", plot = g_Genus, 
       device = "pdf", dpi = 300, width = 18, height = 12, 
       units = "cm")

g_Genus
```

